---
title: "Tidy_model"
author: "Audrielle Staples"
date: "2025-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('tidyverse') ; 
library('modeldata')
``` 


## Introduction

This R Markdown document demonstrates my abilities to use models for data analysis using a data set collected on crickets.


```{r load data, include = FALSE}
data(crickets, package = "modeldata")
names(crickets)
```


## What is that sound?
In this report, we examine what factors predict a cricket's chirp rate.

```{r summary, echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) +
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)') +
  scale_fill_manual(values = c("blue", "red"))
``` 

## Temperature affects chirp rate

```{r temp, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)

summary.lm(temp_lm)
```
Based on a scatter plot of temperature and chirping, it seems that as temperature increases, the rate of chirping also increases. Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute.

## Species-specific effects of temperature on chirping

```{r species, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

species_lm <- lm(rate ~ temp + species, crickets)

summary.lm(species_lm)
```

Using the species-specific factor improves our understanding of the relationship between temperature and chirping. By separating the data points for each species, the plot  shows two  trends, with one species having a higher starting chirp rate than the other. This indicates that the two cricket species respond differently to temperature changes. The slope for temperature is lower (3.6 chirps per minute per degree Celsius) shows that temperature still influences chirping, but the species factor accounts for a portion of the variation. The coefficient for the species variable (−10) indicates that one species, O. exclamationis, chirps 10 times more per minute than the other species, O. niveus, when temperature is held constant. Comparing the R-squared values from the two models shows that the species-specific model explains a greater portion of the variance in chirping rates (99%) compared to the general temperature-only model (92%). This suggests that including species in the model provides a more accurate and reliable prediction of chirping rates.

## Interactions

```{r species historgram, echo = FALSE}
ggplot(crickets, aes(x = rate, fill = species)) +
  geom_histogram(position = 'identity', alpha = 0.7, bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```

The species histogram shows the distribution of temperatures for each cricket species. From this plot, we can see that the two species occupy distinct temperature zones. One species appears to prefer higher temperatures, while the other is found more frequently in cooler temperature ranges. Species have different environmental preferences, which could influence their chirping behaviors.

```{r interactions, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')
  
species_x_temp_lm <- lm(rate ~ temp + species + temp:species, crickets) 
summary.lm(species_x_temp_lm)

anova(species_lm, species_x_temp_lm)
```

After checking for interactions between temperature and species, I decided to stick with the species-specific model. While interactions were present, they didn’t significantly improve the model.
